### 2020-10-24 manuscript/manuscript.rmd
updaed comments from Antonis

---
bibliography: manuscript.bib
csl: nature.csl
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
  word_document: default
  github_document: default
---

@@

Yuanning Li^1,3^, Xing-xing Shen^2,3^, Benjamin Evans^4^, Casey W. Dunn^1^* and Antonis Rokas^3^*

@@

^3^Department of Biological Science, Vanderbilt University, Nashville, TN 37235, USA

@@

There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most difficult problems in animal phylogenetics in part because this is the oldest part of the animal tree. This question has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differed in many respects, including genes considered, species considered, molecular evolution models, and software. Here we systematically explore this question under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing new standardized analyses. It had previously been suggested that site heterogeneous models favor Porifera-sister, but we find that this is not the case. Porifera-sister is only obtained under a narrow set of conditions when the categories are unconstrained in number and include a large number of categories that apply to a small fraction of data. Our analyses help understand the sources of variation across previous studies, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Figs. 1B, 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals, was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Figs. 1A, 1D). Since then, many more relevant phylogenomic studies have been published (Fig. 2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1). The extensive technical variation across these studies has been important to advancing the understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig. 3A). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. This is because methodological variance is often confounded, for example when a particular model is only available in one software package. There are a variety of things that make this a particularly challenging problem. For one, the relevant nodes are the deepest in the animal tree of life. Other factors that have been invoked include concern that Ctenophora-sister could be an artefact of long-branch attraction to outgroups [@pisani2015genomic]. While some sponges have a longer root to tip length, indicating an accelarated rate of evolution in those lineages, the relatively long stem branch of Ctenophora (*e.g.* Extended Data Figs. 1, 2), together with a more typical root-to-tip distance, is instead consistent with a more recent divergence of extant ctenophores than extant poriferans [@podar2001].

To advance this problem it is therefore critical to synthesize published studies and explore variation in methods and results in a standardized and systematic framework. Here, we synthesize data and results from previous phylogenomic analyses that tested Ctenophora-sister and Porifera-sister from 15 studies (Fig. 2A, Extended Data Table 1). These include all such phylogenomic studies of amino acid sequence data published before 2019 for which we could obtain data matrices with gene partition annotations. Among these 15 studies,  three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].

![](figures/Figure_overview.png)
**Fig. 1.** Two alternative phylogenetic hypotheses from previous phylogenomic studies. (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta. (D) A ctenophore, commonly known as comb jellies. (E) A poriferan, commonly known as sponges.

@@

The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $R$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $R$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR). The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), with the same matrix used for all sites. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The published analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. While studies often discuss CAT and WAG models as if they are exclusive, we note that these particular terms apply to non-exclusive model components – CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (*e.g.* GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation. Some analyses partition the data by genes, and use different models for each gene (Supplementary Information section S2).

@@

**Fig. 2.** An overview of previous phylogenomic studies on animals. (A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study. (B) Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of species sampled.

Ingroup taxon sampling also varies widely between studies (Fig. 2B, Extended Data Fig. 4). Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time as studies add additional taxa from these poorly studied groups (Extended Data Fig. 4). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 4). This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were in earlier studies.

@@

Our summary of previous phylogenetic analyses (Fig. 3A) shows that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination. The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) yield different results. This indicates that the traditional framing of the problem, that accommodating rate heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.

### Category number explains differences between heterogeneous analyses

There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (*e.g.* number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here among site heterogeneous models. In published analyses of the animal root, these factors were confounded since all previous heterogeneous analyses used the CAT model in PhyloBayes. Here we seek to deconfound these factors to gain a finer-grained perspective on why results differ between analyses of the same matrices.

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. The standard CAT model employs a Dirichlet process prior to inferring the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from 10 (C10) to 60 (C60). Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses and `r cat_categories$n_categories_gtr  %>% mean(na.rm=TRUE) %>% round(1)` categories for GTR+CAT analyses. This requires a very large number of additional parameters that must be estimated from the data. Though C60 models have a better fit than site-homogeneous models in ModelFinder, the program reported that the C60 model might be overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 is probably already too high according to ModelFinder (Supplementary Table 5). These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].

We examined the specific impact of these large difference in category number on animal root position. It is not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software or other implementation factors, but due to the large difference, in excess of ten-fold, in the number of site categories. When we increased the number of categories in PhyloBayes nCAT analyses, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+nCAT model.

@@

These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017 Extended Data Fig. 1; Philippe2009 Extended Data Fig. 2). Outgroup sampling must be restricted (Choanimalia), a Poisson rather than GTR exchange matrix must be used, and a very large number of site categories must be used (unconstrained CAT, giving hundreds of equilibrium frequency categories). Analyses under other conditions either support Ctenophora-sister or are unresolved. This is consistent across published analyses and our new panels of analyses. The question, then, isn't why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects. If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several things we can learn from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.

There are several impacts of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister). It could be that there are a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categories are available. This would suggest that we should favor the result under a very large number of categories, Porifera-sister. Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameters. This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister. One way to assess this is to look at the allocation of sites to equilibrium frequency categories. Each new category adds the same number of parameters, but that category can be allocated to any number of sites. If as we add categories those categories are allocated to a small fraction of sites, then the cost of those additional categories is high relative to the fraction of the data they help explain, and we should be skeptical of the results when adding so many categories. That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig. 5A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 5A-B). This suggests that support for Porifera-sister is due to adding a large number of parameters that apply to a small fraction of the data. Moreover, we also conducted Bayesian cross-validation experiments employed in PhyloBayes to evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices. Using 10 cross-validation replicates, we found that in half of the replicates, Poisson+nCAT60 models fit these data better than Poisson+CAT models (Extended Data Fig. 6). These results are further corroborated with our *t*-test analysis that there is no significant differences between the model-fit of CAT and nCAT60 models (Whelan2017_strict: *p*-value=`r P_value_Whelan2017_strict`; Philippe2009_Choanimalia: *p*-value=`r P_value_Philippe2009`).

@@

```
![](figures/Figure4_manuscript.png)
**Fig. 4.**  Sensitive analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories. (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values from the sensitive analyses are shown in Supplementary Table 6.

@@

To further explore the variation described above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 7). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrix, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis when using site-homogeneous models in IQ-TREE. Moreover, the outgroup choice has little impact on the distribution of the support for phylogenetic signals in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 7; Supplementary Table 7). In contrast to the C60 models, the phylogenetic signal largely increased in Poisson+CAT models in PhyloBayes towards Porifera-sister, and the outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 7). For example in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).

@@

Resolving the placement of the root in the animal tree of life has been a very difficult phylogenetic problem [@laumer2019revisiting]. By synthesizing past phylogenomic studies and performing new analyses, we confirm that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses. This demonstrates an important interaction between model selection and taxon sampling. Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency. Site-homogeneous models have a single category that is applied to all sites. C models, as implemented in IQ-TREE, have 10-60 categories. CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT. When outgroup sampling is fixed to a restricted subset (a requirement for any analysis to support Porifera-sister), root position is dependent on the number of categories. As the number of categories is increased, at some point support for Ctenophora-sister diminishes and support for Porifera-sister increases (Fig. 4). As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.

>>>>

---
bibliography: manuscript.bib
csl: nature.csl
header-includes:
   - \usepackage{lineno}
   - \linenumbers
output:
  word_document: default
  pdf_document:
    latex_engine: xelatex
  html_document: default
  github_document: default
---

@@

Yuanning Li^1,3^, Xing-Xing Shen^2,3^, Benjamin Evans^4^, Casey W. Dunn^1^* and Antonis Rokas^3^*

@@

^3^Department of Biological Sciences, Vanderbilt University, Nashville, TN 37235, USA

@@

There has been considerable debate about the placement of the root in the animal tree of life, which has emerged as one of the most difficult problems in animal phylogenetics. Resolving the debate has major implications for our understanding of the earliest events in animal evolution, including the origin of the nervous system. Some phylogenetic analyses support a root that places the first split in the phylogeny of living animals between sponges and all other animals (the Porifera-sister hypothesis), and others find support for a split between comb jellies and all other animals (Ctenophora-sister). These analyses differ in many respects, including in the genes considered, species considered, molecular evolution models, and software. Here we systematically explore the rooting of the animal tree of life under consistent conditions by synthesizing data and results from 15 previous phylogenomic studies and performing a comprehensive set of new standardized analyses. It has previously been suggested that site-heterogeneous models favor Porifera-sister, but we find that this is not the case. Rather, Porifera-sister is only obtained under a narrow set of conditions when the categories of sites are unconstrained in number and include a large number of categories that apply to a small fraction of data. Our analyses shed light on the sources of variation between studies aimed at resolving a highly contentious and important branch of the tree of life, and the datasets and analyses consolidated here will be a useful test-platform for development of phylogenomic methods.

@@

Historically, there was little debate about the root of the animal tree of life. Porifera-sister (Fig. 1E), the hypothesis that the animal root marks the divergence of Porifera (sponges) from all other animals (Fig. 1B), was widely accepted though rarely tested. By contrast, there has long been uncertainty about the placement of Ctenophora (comb jellies) (Fig. 1D) in the animal tree of life [@Wallberg:2004ws]. The first phylogenomic study to include ctenophores [@Dunn:2008ky] suggested a new hypothesis, now referred to as Ctenophora-sister, that ctenophores rather than sponges are our most distant living animal relative (Fig. 1A). Since then, many more phylogenomic studies have been published (Fig. 2), with some analyses finding support for Ctenophora-sister, some for Porifera-sister, and some neither [@King:2017ie] (Extended Data Table 1, Supplementary Information section S1). The extensive technical variation across these studies has been important to advancing our understanding of the sensitivity of these analyses, demonstrating for example that outgroup and model selection can have a large impact on these results (Fig. 3A). But the extensive technical variation has also made it difficult to synthesize these results to understand the underlying causes of this sensitivity. Several factors make resolution of the root of the animal tree a particularly challenging problem. For one, the branch in question is the deepest in the animal tree of life. Another factor that has been invoked is that Ctenophora-sister could be an artefact of long-branch attraction to outgroups [@pisani2015genomic]. While some sponges have a longer root to tip length, indicating an accelerated rate of evolution in those lineages, the relatively long stem branch of Ctenophora (*e.g.* Extended Data Figs. 1, 2), together with a more typical root-to-tip distance, is instead consistent with a more recent divergence of extant ctenophores than extant poriferans [@podar2001].

To advance this debate it is critical to examine variation in methods and results in a standardized and systematic framework. Here, we synthesize data and results from 15 previous phylogenomic studies that tested the Ctenophora-sister and Porifera-sister hypotheses (Fig. 2A, Extended Data Table 1). This set includes all phylogenomic studies of amino acid sequence data published before 2019 for which we could obtain data matrices with gene partition annotations. Among these 15 studies, three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].

![](figures/Figure_overview.png)
**Fig. 1.** Two alternative phylogenetic hypotheses for the root of the animal tree. (A) The Ctenophora-sister hypothesis posits that there is a clade (designated by the orange node) that includes all animals except Ctenophora, and that Ctenophora is sister to this clade. (B) The Porifera-sister hypothesis posits that there is a clade (designated by the green node) that includes all animals except Porifera, and that Porifera is sister to this clade. Testing these hypotheses requires evaluating the support for each of these alternative nodes. (C) The animals and their outgroup choice, showing the three progressively more inclusive clades Choanimalia, Holozoa, and Opisthokonta. (D) A ctenophore, commonly known as comb jellies. (E) A poriferan, commonly known as a sponge.

@@

The models of sequence evolution in the studies considered here differ according to two primary components: the exchangeability matrix $R$ and amino acid equilibrium frequencies $\Pi$. The exchangeability matrix $R$ describes the relative rates at which one amino acid changes to another. The studies considered here use exchangeabilities that are the same between all amino acids (Poisson, also referred to as F81),or different; if different, the exchangeabilities can either be fixed based on previously empirically estimated rates (WAG or LG), or independently estimated from the data (GTR). The analyses considered here have site-homogeneous exchangeability models (site-homogeneous model), which means that the same matrix is used for all sites. The equilibrium frequencies describe the expected frequency of each amino acid, which captures the fact that some amino acids are much more common than others. The published analyses differ in whether they take a homogeneous approach and jointly estimate the same frequency across all sites in a partition, or add parameters that allow heterogeneous equilibrium frequencies that differ across sites. Heterogeneous approaches include CAT [@Lartillot:2004dq], which is implemented in the software PhyloBayes and has been widely applied to questions of deep animal phylogeny. The models that are applied in practice are heavily influenced by computational costs, model availability in software, and convention. While studies often discuss CAT and WAG models as if they are mutually exclusive, we note that these particular terms apply to non-exclusive model components – CAT refers to heterogeneous equilibrium frequencies across sites and WAG to a particular exchangeability matrix. In this literature, CAT is generally shorthand for Poisson+CAT and WAG is shorthand for WAG+homogeneous equilibrium frequency estimation. To avoid confusion on this point, here we always specify the exchangeability matrix first (*e.g.* GTR), followed by modifiers that describe the accommodation of heterogeneity in equilibrium frequencies (*e.g.* CAT). If site homogeneous equilibrium frequencies are used, we refer to the exchangeability matrix alone. Gamma-rate heterogeneity, a scalar that accommodates total rate of change across sites, is used in almost every analysis conducted here and we generally omit its designation. Some analyses partition the data by genes, and use different models for each gene (Supplementary Information section S2).

@@

**Fig. 2.** An overview of previous phylogenomic studies on animals. (A) A time-series showing the main conclusions of previous phylogenomic studies that indicated their analyses supported Ctenophora-sister (orange nodes) or Porifera-sister (green nodes) based on their main conclusion from each study. (B) Each of the primary matrices considered here, color coded by taxon sampling. Horizontal size is proportional to the number of genes sampled, vertical size to the number of species sampled. Only 11 matrices were shown here since three studies [@pisani2015genomic; @whelan2016let; @feuda2017improved] are based entirely on previously published data and gene-partition data is not available from one study [@Pick:2010eb].

Ingroup taxon sampling also varies widely between studies (Fig. 2B, Extended Data Fig. 4). Sampling of ingroup taxa (animals) in early studies was biased toward Bilateria. Sampling of non-bilaterian animals, including sponges and ctenophores, has improved over time (Extended Data Fig. 4). Within each clade, there is often considerable variation in taxon sampling and therefore often little species overlap across studies (Extended Data Fig. 4). This variation is in part because newer sequencing technologies in more recent studies are usually not applied to the exact same species that were included in earlier studies.

@@

Our summary of previous phylogenetic analyses (Fig. 3A) shows that Ctenophora-sister is supported in analyses that span the full range of outgroup sampling and models used to date, including some analyses with restricted outgroup sampling and models that accommodate site-heterogeneous equilibrium frequencies with CAT. This is consistent with previous assessments of the problem [@whelan2016let], but is drawn from a much more extensive and systematic examination. The only analyses that support Porifera-sister have a reduced outgroup sampling (Choanimalia, Holozoa) and site-heterogeneous models with CAT. Model adequacy assessments generally favor GTR+CAT over Poisson+CAT or site-homogeneous models [@pisani2015genomic; @feuda2017improved], but because GTR+CAT is so parameter rich, many analyses that use a model with GTR+CAT do not converge. The fact that Porifera-sister is recovered only for particular models with particular outgroup sampling indicates that model and outgroup interact, and that this interaction is fundamental to understanding the range of results obtained across analyses.

@@

Our new analyses show that, with restricted outgroup sampling, analyses of the same matrices with two different means of accommodating site heterogeneity in equilibrium frequencies (C60 in IQ-TREE and CAT in PhyloBayes) yield different results. This indicates that the traditional framing of the problem, that accommodating site heterogeneity leads to support for Porifera-sister, is not correct. Instead, there is something about PhyloBayes CAT analyses specifically that leads to support for Porifera-sister.

### Category number explains differences between site-heterogeneous analyses

There are several factors, including variations in models (C60 vs CAT), software (Phylobayes vs IQ-TREE), and implementation details (*e.g.* number of categories used to accommodate site heterogeneity) that could explain the new variation in results noted here among site-heterogeneous models. In published analyses of the animal root, these factors were confounded since all previous heterogeneous analyses used the CAT model in PhyloBayes. Here we seek to deconfound these factors to gain a finer-grained perspective on why results differ between analyses of the same matrices.

@@

A primary difference between the C (like C60) and CAT site-heterogeneous models is the number of equilibrium frequency categories. The standard CAT model employs a Dirichlet process prior to inferring the number of equilibrium frequency categories, so the number of categories is variable [@Lartillot:2004dq]. IQ-TREE implements C models [@si2008empirical] with a fixed number of categories, from 10 (C10) to 60 (C60). Differences in analysis results could therefore be due to differences in the number of categories. The number of categories inferred by CAT in PhyloBayes can be very high (Supplementary Table 4), with a mean here of `r cat_categories$n_categories_poisson %>% mean() %>% round(1)` categories for Poisson+CAT analyses in all matrices and `r cat_categories$n_categories_gtr  %>% mean(na.rm=TRUE) %>% round(1)` categories for GTR+CAT analyses in several representative matrices. This requires a very large number of additional parameters that must be estimated from the data. Though C60 models have a better fit than site-homogeneous models in ModelFinder, the program reported that the C60 model is overfitting because some category frequencies are estimated close to zero (except the Whelan2015, Whelan2017 and Simion2017 matrices), suggesting that 60 categories are probably already too high according (Supplementary Table 5). These findings are consistent with earlier concern that CAT models can overestimate the number of equilibrium frequency categories in simulated matrices [@whelan2016let].

We examined the specific impact of this large difference in category number on animal root position. It is currently not possible to use more than 60 categories for C models in IQ-TREE, but the number of categories can be set *a priori* in Phylobayes using nCAT. We therefore varied the number of categories in Phylobayes analyses (Fig. 4; Supplementary Table 6). We found that Poisson+nCAT60 analyses in PhyloBayes, like Poisson+C60 and WAG+C60 IQ-TREE analyses, provide strong support of Ctenophora-sister. This indicates that the difference in results between unconstrained CAT analyses in PhyloBayes and C60 analyses in IQ-TREE is not due to differences in the software or other implementation factors, but due to the large difference, in excess of ten-fold, in the number of site categories. When we increased the number of categories in PhyloBayes nCAT analyses, we observe the transition from support for Ctenophora-sister to an unresolved root to Porifera-sister (Fig. 4). For example, for the Whelan2017_strict matrix this transition occurs between 60 -- 120 categories when using Poisson+nCAT model.

@@

These results further clarify when analyses support Porifera-sister (*e.g.* Whelan2017: Extended Data Fig. 1; Philippe2009: Extended Data Fig. 2): when outgroup sampling is restricted (Choanimalia), when a Poisson (rather than a GTR) exchange matrix is used, and when a very large number of site categories is used (unconstrained CAT, giving hundreds of equilibrium frequency categories). Analyses under other conditions either support Ctenophora-sister or are unresolved. This is consistent across published analyses and our new panels of analyses. The question, then, isn’t why similar analyses give different results, but how we should interpret variation in results when we run analyses that differ in these specific respects. If we fix the first two features to conditions that are necessary for Porifera-sister support (Choanimalia taxon sampling and a Poisson exchange matrix), there are several insights that we can glean from examining how the number of equilibrium frequency categories impacts results that sheds light on interpretation of those results.

There are several effects of increasing the number of equilibrium frequency categories from dozens (which leads to support for Ctenophora-Sister) to hundreds (which supports Porifera-sister). It could be that there is a large fraction of sites with diverse equilibrium frequencies that are not accommodated well when there is a smaller number of categories, and that by increasing the number of categories the model is able to overcome biases when too few categories are available. This would suggest that we should favor the result under a very large number of categories, Porifera-sister. Alternatively, it could be that adding a large number of categories is adding so many parameters that the information in the data are insufficient to infer both the topology and those parameters. This would suggest that we should favor the result under a smaller number of categories, Ctenophora-sister. One way to assess this is to look at the allocation of sites to equilibrium frequency categories. Each new category adds the same number of parameters, but that category can be allocated to any number of sites. If as we add categories those categories are allocated to a small fraction of sites, then the cost of those additional categories is high relative to the fraction of the data they help explain. That is exactly what we see when we examine category allocations in the last chain samples of analyses of the Philippe2009_Choanimalia and Whelan2017_strict matrices (Extended Data Fig. 5A-B), which have `r n_categories_phil_cat` and `r n_categories_whel_cat` categories, repectively. The fraction of sites allocated to the 50% of the least frequent categories is `r mid_phil_cat`% in the analysis of the Philippe2009_Choanimalia matrix and `r mid_whel_cat`% in the analysis of the Whelan2017_strict matrix. This long tail of categories that apply to a very small fraction of sites is in stark contrast to nCAT60 analyses, constrained to 60 categories, which have no such long tail of rare categories (Extended Data Fig. 5A-B). This suggests that support for Porifera-sister is due to adding a large number of parameters that apply to a small fraction of the data. To formally evaluate the predictive performance and model-fit between Poisson+CAT and Poisson+nCAT60 models in Whelan2017_strict and Philippe2009_Choanimalia matrices, we conducted Bayesian cross-validation experiments employed in PhyloBayes. Using 10 cross-validation replicates, we found that in half of the replicates, Poisson+nCAT60 models fit these data better than Poisson+CAT models (Extended Data Fig. 6); a t-test analysis shows that there is no significant difference between the model-fit of CAT and nCAT60 models (Whelan2017_strict: *p*-value=`r P_value_Whelan2017_strict`; Philippe2009_Choanimalia: *p*-value=`r P_value_Philippe2009`).

@@

```
![](figures/Figure4_manuscript.png)
**Fig. 4.**  Sensitivity analyses with representative data matrices analyzed by different number of equilibrium frequency categories (nCAT) in PhyloBayes. Statistical support values (posterior probabilities) were obtained from three data matrices using the site-heterogeneous Poisson+CAT model with different categories. (A) Phlippe2009_Choanozoa; (B) Philippe2009_Holozoa; (C) Whelan2017_full_Choanozoa; (D) Whelan2017_strict. Statistical support for Ctenophora-sister and Porifera-sister is indicated in orange and green, respectively. Support values from the sensitive analyses are shown in Supplementary Table 6.

@@

To further explore the phylogenetic signal of different models we discussed above, we quantified the phylogenetic signal for Porifera-sister and Ctenophora-sister topologies across three representative data matrices when varying outgroup sampling and model (Extended Data Fig. 7). By calculating differences in log-likelihood scores for these topologies for every gene ($\Delta$|lnL|) in each matrixin each matrix when using site-homogeneous models in IQ-TREE, we found that the Ctenophora-sister had the higher proportions of supporting genes in every analysis. Moreover, outgroup choice has little impact on the distribution of the support for phylogenetic signals in analyses with site-homogeneous models. This finding is largely consistent with the previously observed distribution of support for Ctenophora-sister in other data matrices [@shen2017contentious].

Although a higher proportion of genes support Ctenophora-sister with site-heterogeneous C60 models, the phylogenetic signal decreases in many genes using C60 models compared to site-homogeneous models. In an extreme case, in matrices from Ryan2013_est nearly 30% of genes changed from strong $\Delta$|lnL|>2) to weak Ctenophora-sister signal ($\Delta$|lnL|<2) (Extended Data Fig. 7; Supplementary Table 7). In contrast to the C60 models, there is a major increase in phylogenetic signal in Poisson+CAT models in PhyloBayes towards Porifera-sister, and outgroup choice has a major effect of the distribution of phylogenetic signal (Extended Data Fig. 7). For example, in Whelan2017_full matrix, we found that the number of genes that support Ctenophora-sister in analyses with CAT decreases from 57.5% in matrices with distant outgroups (Holozoa) to 35.4% when outgroups are restricted (Choanimalia, Supplementary Table 7).

@@

Resolving the placement of the root in the animal tree of life has proved very challenging [@laumer2019revisiting]. By synthesizing past phylogenomic studies and performing new analyses, we find that support of Porifera-sister is only recovered by site-heterogeneous CAT models with restricted outgroup sampling, and then only in some of such analyses. This demonstrates an important interaction between model selection and taxon sampling. Through controlled analyses we are able to identify the specific aspect of the models that is involved in this variation -- the number of categories used to accommodate site heterogeneity in equilibrium frequency. Site-homogeneous models have a single category that is applied to all sites. C models, as implemented in IQ-TREE, have 10-60 categories. CAT models can estimate the number of categories, in which case they use more than 500 for the matrices assessed here, or set to a specific number with nCAT. When outgroup sampling is fixed to a restricted subset (a requirement for any analysis to support Porifera-sister), root position is dependent on the number of categories. As the number of categories is increased, at some point support for Ctenophora-sister diminishes and support for Porifera-sister increases (Fig. 4). As part of this process, the number of genes with phylogenetic signal for each hypothesis also changes.

Overcoming the challenges associated with resolving this difficult and important branch of the tree of life goes beyond models. For example, there are many organisms that are highly relevant to this problem, in particular outgroup, ctenophore, and sponge taxa, for which no genome or transcriptome data are available [@King:2017ie]. More broadly, there are very few chromosome-level genome assemblies for animals outside of Bilateria. But resolution of the debate will likely require more than just more data – for example, high quality chromosome-level assemblies will greatly facilitate gene homology assessments made during matrix construction, reduce the possibility of contamination, and address occupancy issues in transcriptomic data. We hope that the work we have conducted here to consolidate many datasets and analyses in standard formats will make it easier for other investigators to engage in this particularly interesting and difficult phylogenetic problem, and also to develop methods and tools that will help with other difficult phylogenetic problems.
